---
title: "Living Data Project Manuscript"
subtitle: "***Effects of Brook Trout on Juvenile Chinook Salmon Survival***"
author: "Sophia Kast^1^*"
date: "`r Sys.Date()`"
# format for Journal of Ecology
geometry: margin=1in
fig_caption: true
header-includes: 
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{fontspec} # can only be used with xelatex output
  - \setmainfont{Times New Roman} # can only be used with xelatex output
  
output: 
  pdf_document:
      latex_engine: xelatex


csl: 'journal-of-ecology.csl'
bibliography: ['Literature.bib', 'grateful-refs.bib']
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1.  Biology Department, University of British Columbia, Okanagan Campus

\*Corresponding Author

\newpage

# Abstract

Invasive species can affect the survival of indigenous species. In this project, I use data from a 2002 study looking at the effect of brook trout on the survival of juvenile Chinook salmon in the Salmon River watershed in Idaho, in the western United States. I visualize the data to see if there is a correlation between presence of brook trout and survival of juvenile salmon. The overlying purpose of this project is to learn best practices for open science workflows and understand how to create reproducible scientific research projects from start to finish.

**Key-words:** brook trout, Chinook salmon, salmon survival

# Introduction

As the planet becomes increasingly spatially connected by humans, there are increasing conduits for other species to move out of their native ranges. This can have a variety of effects. Sometimes, novel species in a region can detrimentally affect the survival of established indigenous populations. Brook trout, a salmonid fish native to northeastern North America. Its range has artificially expanded and it is now one of the most populous non-native fish species in the western United States. It has been suspected that these trout may negatively affect native salmon populations in western watersheds. In this project, I borrow data from a 2002 study investigating juvenile Chinook salmon survival in the Salmon River watershed, where some streams have robust brook trout populations and others do not. At each site, researchers tagged juvenile salmon in the fall. The following spring, these tagged salmon were tracked at the Lower Granite Dam to determine the number of survivors [@levin2002].

The purpose of this mini-project is to demonstrate knowledge of open science practices and workflows. To do this, I will simply create 3 different plots of the data associated with the 2002 study by Levin et al. [@levin2002]. 

# Methods

First, set up the project directory: 

```{r create output directories if needed, echo=TRUE}
if( ! dir.exists("data") ){ dir.create("data") }
```

Load packages using ```groundhog```. The package ```grateful``` is used to cite packages, but cannot be loaded using ```groundhog``` due to its storage location on GitHub. 

```{r loadlib, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library('grateful')
library('groundhog')
groundhog.library(c('tidyverse', 'tinytex'), '2022-09-01')
```

Run the function 
```
get_pkgs_info()
```
This creates the package bibliography. The output is not shown here as it is not necessary for the purposes of this manuscript. 


```{r citepkgs, message=FALSE, warning=FALSE, include=FALSE}
get_pkgs_info()
```

Read the data stored at the following url: <https://whitlockschluter3e.zoology.ubc.ca/Data/chapter12/chap12e4ChinookWithBrookTrout.csv>. 

```{r loaddata, echo=TRUE, message=FALSE, warning=FALSE}
salmon_raw <- readr::read_csv('https://whitlockschluter3e.zoology.ubc.ca/Data/chapter12/chap12e4ChinookWithBrookTrout.csv')
```

Tidy the data by changing the column names. Add a column to indicate site location number. Do not conduct calculations. The proportion of salmon surviving at Lower Granite Dam is already present in the raw table. 

```{r tidydata, echo=TRUE, message=FALSE, warning=FALSE}
salmon <- mutate(salmon_raw,
                 troutTreatment = stringr::str_to_title(troutTreatment),
                 site = row_number()) %>% 
  rename(brook_trout_presence = troutTreatment, 
       salmon_released = nReleased,
       surviving_salmon = nSurvivors,
       proportion_survived = proportionSurvived) %>%
  relocate(site, .before = brook_trout_presence)
```

Write the file to a .csv to reference when creating plots.  

```{r echo=TRUE}
write.csv(salmon, file = 'data/salmon_clean.csv')
```

\newpage

# Results

Visualize the data in 3 separate plots to gain experience creating plots in RStudio. 

First, create a scatterplot showing the number of surviving salmon based on the number of released salmon. 


```{r scatter, echo=FALSE, warning=TRUE, fig.cap = 'Salmon Survival at Lower Granite Dam'}

plot(salmon$surviving_salmon~salmon$salmon_released, 
     xlab = '# of Salmon Released', 
     ylab = '# of Salmon Surviving',
     col = "black")
```

\newpage

Next, create a strip chart showing the proportion of surviving salmon for each brook trout treatment (presence or absence). 


```{r stripchart, echo=FALSE, fig.cap = 'Salmon survival related to brook trout presence as a strip chart'}

salmon %>%
  ggplot(aes(x = brook_trout_presence, y = proportion_survived)) +
  geom_jitter(colour = "black", size = 3, shape = 1, width = 0.1) +
  xlab("Brook Trout Population") +
  ylab("Proportion of surviving salmon") +
  ylim(0, 0.5) +
  theme_bw()
```

\newpage

Finally, create a violin chart to represent the same data as in the stripchart, but with a different visual style. 


```{r violin, echo=FALSE, fig.cap = 'Salmon survival related to brook trout presence as a violin plot'}

surviving_salmon.violin <- salmon %>% 
  ggplot(aes(x = brook_trout_presence, y = proportion_survived)) +
  geom_violin() +
  xlab("Brook Trout Population") +
  ylab("Proportion of Surviving Salmon") +
  theme_bw()
surviving_salmon.violin.points <- surviving_salmon.violin + geom_jitter(colour = "black", size = 3, shape = 1, width = 0.1)
surviving_salmon.violin.points
```


Finally, run the following function to cite the packages in the manuscript. 

```{r echo=TRUE, results='asis'}
cite_packages(output = 'paragraph')
```

# Discussion

The results of this project show  that brook trout may be affecting the survival of juvenile Chinook salmon as they make their way from spawning grounds to the Lower Granite Dam. Results from 3 of the sites where brook trout were absent shower much higher survival rates of salmon. However, 3 other sites without brook trout showed no difference in survival rate from those with brook trout. This could indicate that there were other issues impacting brook trout survival. These factors could include increased water temperatures, pressure from sport fishing, other predation pressures, water pollution, and others. Further research is necessary to understand the exact effects that brook trout may have on Chinook salmon survival in the Salmon River watershed. 

\newpage

# Acknowledgements

Thanks to the LDP course professor, instructors, and TA for guiding me through this process. 

# References

::: {#refs}
:::


